Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-10-31 analysis/paper/article.qmd
text in background section

@@

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically contributors copy and paste the text of job ads from other sources, such as the _Chronicle of Higher Education_, _Higher Ed Jobs_, and university websites, into the wiki, collecting ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position, such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

@@

Our results reveal some of these shifts, providing insights into the dynamics of topical, methods, and geographic foci over time. While these dynamics are deeply felt by people actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale. Trends in job ads reflect broader shifts in intellectual and practical priorities about archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. The downward trend in tenure-track positions during 2013-2019 related to decline in undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities adapted to online instruction and assessment in an effort to minimize the spread of the virus. 
In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities tailored the content of the job ads to focus on topics where Black and Indigenous researchers are most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

@@

A key limitation of our research is that it does not include an analysis of the academic profiles of those who were eventually hired from the job ads. Our results reveal collective aspirations for the future of the field, but do not show how these worked out in the topical and geographic foci of the people who were eventually hired.

>>>>

The academic job market is a source of uncertainty, both for job seekers, as they look for ads that are well-matched to their niche of skills and experience that have taken years to cultivate and refine, and for hiring committees as they sift through sometimes hundreds of applicants in the hope of finding a candidate that meets all of their needs. For many academics there is a perception that the last two to three decades have seen the job market become increasingly difficult for job seekers. In many fields it is common to hear that there are fewer jobs available relative to the number of PhD graduates, higher numbers of short-term temporary positions relative to secure tenure-track positions, and increasingly complex application requirements with specific statements (e.g. teaching, research, and diversity) that need to be tailed for each application, making the process of applying for jobs a full-time job of its own.

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echos findings from other fields, for example in biomedical sciences where there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from 2021–22 Faculty indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Research into

abysmal

@@

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically contributors copy and paste the text of job ads from other sources, such as the _Chronicle of Higher Education_, _Higher Ed Jobs_, and university websites, into the wiki, collecting ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position, such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

@@

Our results reveal some of these shifts, providing insights into the dynamics of topical, methods, and geographic foci over time. While these dynamics are deeply felt by people actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale. Trends in job ads reflect broader shifts in intellectual and practical priorities about archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. The downward trend in tenure-track positions during 2013-2019 related to decline in undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus.

In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities tailored the content of the job ads to focus on topics where Black and Indigenous researchers are most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

@@

A key limitation of our research is that it does not include an analysis of the academic profiles of those who were eventually hired from the job ads. Our results reveal collective aspirations for the future of the field, but only from those who are securely employed in it as tenured professors serving on hiring committees at US universities. Our results do not show how these aspirations worked out in the topical and geographic foci of the people who were eventually hired for the positions studied here. Future work should consider interviewing faculty hired during our study period to match up people to the ads they responded to. If the successful applicants can be identified, then the match between the details of the job ad and the applicant's research can be analysed. Then we can assess how effective the hiring process is for driving change in the discipline by setting the topics of courses that will be taught in undergraduate and graduate curricula, and research that will be supported by universities.
